---
title: "EDS 223: Homework 4"
subtitle: "TBD"
author: Stephen Carroll
date: 12-14-24
execute: 
  eval: true
  warning: false
  message: false
  echo: true
format:
  html:
    toc: true
    code-fold: true
editor_options: 
  chunk_output_type: console
---

# Prioritizing potential aquaculture


## Background: 

The West Coast of the United States has five Exclusive Economic Zones (EEZ). An EEZ is a maritime zone that extends up to 200 nautical miles (about 370 kilometers) from a country's coastline. These zones can be analyzed to determine suitable locations for marine aquaculture. By using geospatial data on sea surface temperature (SST) and bathymetry (depth), we can identify areas within these zones that meet the environmental requirements for aquaculture of individual species. This project aims to develop a map of the suitable habitat areas for Oysters. We will also build a function that will be capable of assessing potential sites for any species based on preferred SST and depth, and creating a map for them, contributing to the sustainable growth of marine aquaculture.

## Setup
```{r}
# Load libraries
library(sf)
library(here)
library(terra)
library(tmap)
library(tidyverse)
library(viridisLite)
library(RColorBrewer)
library(kableExtra)
library(testthat)
library(patchwork)
library(testthat)
```

## Load data

```{r}
# Read in data 
west_coast_eez <- read_sf(here("data", "wc_regions_clean.shp"))

depth <- terra::rast(here("data", "depth.tif"))
```


## Prepare data

```{r}
# Compile all our SST rasters into one object
sst_files <- list.files(here("data"), 
                        pattern = "average", 
                        full.names = TRUE)

# Read in sst data and store as a raster stack
sst_stack <- terra::rast(sst_files)

# Add labels to the raster layers
names(sst_stack) <- c("sst2008", "sst2009", "sst2010", "sst2011", "sst2012")

# # Preliminary plot
# plot(sst_stack)

```

## CRS check

```{r}
# Confirm that the two raster datasets have the same CRS
if(crs(depth) == crs(sst_stack)) { # Raster data  vs raster data
  print("Coordinate reference systems match.")
} else{
  warning("Updating 'sst_stack' coordinate reference system to match 'depth.'")
  sst_stack <- project(sst_stack, crs(depth))  # Project data to match
}

# Confirm that the vector dataset has the same CRS as the raster dataset
if(st_crs(west_coast_eez) == crs(depth)) { # Vector data vs raster data
  print("Coordinate reference systems match")
} else{
  warning("Updating 'west_coast_eez' coordinate reference system to match 'depth'.")
  west_coast_eez <- st_transform(west_coast_eez, st_crs(depth)) # Transform data to match
}
```

# Test the CRS conversion
```{r}
# Confirm that any conversions were successful with a test
test_that("CRS of all objects are the same", {
  
# Check if CRS of sst_stack matches depth
expect_identical(crs(depth), crs(sst_stack), 
                   info = "CRS of 'sst_stack' does not match 'depth'.")
  
# Check if CRS of west_coast_eez matches depth
expect_identical(st_crs(depth), st_crs(west_coast_eez), 
                   info = "CRS of 'west_coast_eez' does not match 'depth'.")
  
})
```

## Process data

### Make one raster that is comprised of an average of SST from 2008-2012 and convert from Kelvin to Celsius

```{r}
# Find average temperature from 2008-2012
mean_sst <- app(sst_stack, fun = mean)

# Convert units from Kelvin to Celsius
mean_sst <- mean_sst - 273.15
```


### Remove bathymetry data that does not lie in the same extent as the temperature data

```{r}
# Crop depth to match the geographic extent of sst
depth_crop <- crop(depth, mean_sst)

# Re-sample depth to match the resolution of sst
depth_resample <- resample(depth_crop, y = mean_sst, method = "near")
```

### Check that the depth and SST rasters match in resolution, extent, and coordinate reference system

```{r}
# Stack the two rasters to confirm they are compatible
depth_sst_stack <- c(depth_resample, mean_sst)

# Confirm they have the same resolution, extent, and CRS
if(all(terra::res(depth_resample) == terra::res(mean_sst))) {
  if(terra::crs(depth_resample) == terra::crs(mean_sst)) {
    if(terra::ext(depth_resample) == terra::ext(mean_sst)){
      print("All match")
    }else{
      stop("extenet does not match")
    }
  }else{
    stop("CRS does not match")
  }
}else{
  stop("Resolution does not match")
}
```

## Find suitable locations


#### Oysters

Research has shown that oysters need the following conditions for optimal growth:

- Sea surface temperature: 11-30°C
- Depth: 0-70 meters below sea level

### Reclassify SST and depth data into locations that are suitable for oysters

```{r}
# Create sst reclassification matrix
rcl_sst <- matrix(c(-Inf, 11, NA, # min temperature
                    11, 30, 1, 
                    30, Inf, NA), # max temperature
                    ncol = 3, byrow = TRUE)

# Use reclassification matrix to reclassify sst raster
reclass_sst <- classify(mean_sst, rcl = rcl_sst)

# Create depth reclassification matrix
rcl_depth <- matrix(c(-Inf, -70, NA, # min depth
                    -70, 0, 1, 
                    0, Inf, NA), # max depth
                    ncol = 3, byrow = TRUE)

# Use reclassification matrix to reclassify depth raster
reclass_depth <- classify(depth_resample, rcl = rcl_depth)
```

### Find locations that satisfy both SST and depth conditions for oysters

```{r}
# Multiply reclassified matrices to find cells that each have a 1
suitable_sst_depth <- lapp(c(reclass_sst, reclass_depth), fun = "*") # Use multiply function. Only cells that both have a 1 will remain a 1

```

## Determine the most suitable EEZ for oysters by area

```{r}
# Create a raster of the eez data
west_coast_eez_raster <- rasterize(west_coast_eez, 
                                   suitable_sst_depth, 
                                   field = "rgn") # Transfer the region variable 

```

### Find the area of grid cells & the total suitable area within each EEZ 

```{r}
# Quantify area of grid cells suitable for oysters
suitable_area <- cellSize(suitable_sst_depth, # Area that satisfies depth and temperature parameters
                          mask = TRUE, # Keep NAs in output
                          unit = "km") # Match units to eez units

# Sum suitable area by eez region
suitable_eez_area <- zonal(suitable_area, 
      west_coast_eez_raster,
      fun = "sum",
      na.rm = TRUE) 

# Display values in a table
kable(suitable_eez_area, digits = 2, 
        col.names = c("Region","Suitable Area km^2"), # Use LaTex to render the exponent as superscript
        caption = "Suitable Area by Exclusive Economic Zone Region") %>% 
kable_styling(bootstrap_options = "striped",
                position = "center")

```

## Visualize the results

```{r}
# Combine the eez suitable area data with our original eez dataframe
eez_join <- left_join(west_coast_eez, suitable_eez_area, by = "rgn")

# Create a color palette with colorbrewer for bathymetrry data
my_palette2 <- rev(c('#693829FF', '#894B33FF', '#A56A3EFF', '#CFB267FF', '#D9C5B6FF', '#9CA9BAFF', '#5480B5FF', '#3D619DFF', '#405A95FF', '#345084FF'))


tm_shape(depth_crop) +
  tm_raster(palette = my_palette2,
            title = "Bathymetry\n(Meters above or\n below sea level)",
            alpha = 1,
            stretch.palette = FALSE,
            midpoint = 0, #### SHOULD THIS BE NA??? ####
            legend.show = TRUE) +
  tm_shape(eez_join, 
           raster.downsample = TRUE) + # Re-sample to lower resolution
  tm_polygons(col = "area",
              palette = "Greens",
              alpha = 0.75,
              title = expression("Suitable habitat area (km"^2*")")) + # Write units as an expression
  tm_text("rgn",
          xmod = -0.2,
          size = 0.6,) + # Label each region
  tm_compass(size = 1.5,
             position = c("left", "bottom")) +
  tm_scale_bar(position = c("left", "bottom")) +
  tm_layout(legend.outside = TRUE,
            frame = TRUE,
            main.title = "Suitable Oyster Habitat in\nWest Coast Exclusive Economic Zones",
            main.title.size = 1.35,
            main.title.position = c("center", "top"),
            inner.margins = c(0, 0, 0, 0)) # Fit map to extent of frame
```


## Generalize the workflow with a function

```{r}
aquaculture_fun <- function(species, min_sst, max_sst, min_depth, max_depth){
  
  # Read in EEZ and Depth data
  west_coast_eez <- read_sf(here("data", "wc_regions_clean.shp"))
  depth <- terra::rast(here("data", "depth.tif"))
  
  # Create list of sst files
  sst_files <- list.files(here("data"), 
                          pattern = "average", 
                          full.names = TRUE)
  
  # Read in sst data and store as a raster stack
  sst_stack <- terra::rast(sst_files)
  
  # Reproject CRS to EPSG:4326
  depth <- project(depth, crs("EPSG:4326"))
  sst_stack <- project(sst_stack, crs("EPSG:4326"))
  west_coast_eez <- st_transform(west_coast_eez, crs = crs("EPSG:4326"))
  
  # Find average temperature from 2008-2012 and convert to kelvin
  mean_sst <- app(sst_stack, fun = mean)
  mean_sst <- mean_sst - 273.15
  
  # Crop depth to match the geographic extent of sst
  depth_crop <- crop(depth, sst_stack)
  
  # Re-sample depth to match the resolution of sst
  depth_resample <- resample(depth_crop, y = mean_sst, method = "near")
  
  # Create sst reclassification matrix
  rcl_sst <- matrix(c(-Inf, min_sst, NA,
                      min_sst, max_sst, 1,
                      max_sst, Inf, NA),
                    ncol = 3, byrow = TRUE)
  
  # Use reclassification matrix to reclassify sst raster
  reclass_sst <- classify(mean_sst, rcl = rcl_sst)
  
  # Create depth reclassification matrix
  rcl_depth <- matrix(c(-Inf, min_depth, NA,
                        min_depth, max_depth, 1,
                        max_depth, Inf, NA),
                      ncol = 3, byrow = TRUE)
  
  # Use reclassification matrix to reclassify depth raster
  reclass_depth <- classify(depth_resample, rcl = rcl_depth)
  
  # Find locations that satisfy both SST and depth conditions
  suitable_sst_depth <- lapp(c(reclass_sst, reclass_depth), fun = "*") # Use multiply function. Only cells that both have a 1 will remain a 1
  
  # Create a raster of the eez data
  west_coast_eez_raster <- rasterize(west_coast_eez, 
                                     suitable_sst_depth, 
                                     field = "rgn") # Transfer the region variable 
  
  # Quantify area of grid cells suitable for oysters
  suitable_area <- cellSize(suitable_sst_depth, # Area that satisfies depth and temperature parameters
                            mask = TRUE, # Keep NAs in output
                            unit = "km") # Match units to eez units
  
  # Sum suitable area by eez region
  suitable_eez_area <- zonal(suitable_area, 
                             west_coast_eez_raster,
                             fun = "sum",
                             na.rm = TRUE) 
  
  # Combine the eez suitable area data with our original eez dataframe
  eez_join <- left_join(west_coast_eez, suitable_eez_area, by = "rgn")
  
  # Create a color palette with colorbrewer for bathymetrry data
  my_palette2 <- rev(c('#693829FF', '#894B33FF', '#A56A3EFF', '#CFB267FF', '#D9C5B6FF', '#9CA9BAFF', '#5480B5FF', '#3D619DFF', '#405A95FF', '#345084FF'))
  
  
  tm_shape(depth_crop) +
    tm_raster(palette = my_palette2,
              title = "Bathymetry\n(Meters above or\n below sea level)",
              alpha = 1,
              stretch.palette = FALSE,
              midpoint = 0, 
              legend.show = TRUE) +
    tm_shape(eez_join, 
             raster.downsample = TRUE) + # Re-sample to lower resolution
    tm_polygons(col = "area",
                palette = "Greens",
                alpha = 0.75,
                title = expression("Suitable habitat area (km"^2*")")) + # Write units as an expression
    tm_text("rgn",
            xmod = -0.2,
            size = 0.6,) + # Label each region
    tm_compass(size = 1.5,
               position = c("left", "bottom")) +
    tm_scale_bar(position = c("left", "bottom")) +
    tm_layout(legend.outside = TRUE,
              frame = TRUE,
              main.title = paste("Suitable", species, "Habitat in\nWest Coast Exclusive Economic Zones"), # Print species name in title
              main.title.size = 1.35,
              main.title.position = c("center", "top"),
              inner.margins = c(0, 0, 0, 0)) # Fit map to extent of frame
  
}

```

### Apply the function to new species

The parameters for the function are species, minimum and maximum sea surface temperatures, and minimum and maximum depths for the species. The resulting output will be a map indicating the total suitable area for that species within each EEZ.


#### Dungeness Crab

Research has shown that dungeness crabs need the following conditions for optimal growth:

- Sea surface temperature: 3°C - 19°C, preferred 9°C
- Depth: 0 - 360 meters below sea level

```{r}
aquaculture_fun("Dungeness Crab", 3, 19, 0, 360)
```

#### Razor Clam

Research has shown that razor clams need the following conditions for optimal growth:

- Sea surface temperature: 5.6°C - 12.6°C, mean 8.9°C
- Depth: 0 - 55 meters below sea level


```{r}
aquaculture_fun("Razor Clam", 5.6, 12.6, 0, 55)
```

## Reflection

Our function produces a clear representation of the 5 exclusive economic zones on the West Coast of the United States colored by the total suitable habitat area (km$^2$) for the species provided. Central California has by far the most suitable habitat for oysters, followed by Southern California, and Washington. Dungeness crabs are best suited for the Washington, Southern California, adn Northern California EEZs, respectively. Washington and Northern California appear to be the only EEZs with significant habitable area for razor clams.

By creating a function, we were able to easily use all of the steps followed investigating oyster habitat, and use them to generalize a workflow for all species.  This function can now be used to assist scientists and policymakers to understand where they should be directing their aquaculture development efforts on the West Coast.

```{r}
# Create a table for my citations
citations_df = tibble::tribble(
  ~Data,
  ~Citation,
  ~Accesssed,
  ~Link,
 "Sea Surface Temperature Data", "NOAA Coral Reef Watch. 2019, updated daily. NOAA Coral Reef Watch Version 3.1 Daily 5km Satellite Regional Virtual Station Time Series Data. College Park, Maryland, USA: NOAA Coral Reef Watch.", "11/29/24", "[NOAA Coral Reef Watch](https://coralreefwatch.noaa.gov/product/5km/index_5km_ssta.php)",
  "Bathymetry Data", "GEBCO Compilation Group (2022) GEBCO_2022 Grid (doi:10.5285/e0f0bb80-ab44-2739-e053-6c86abc0289c)", "11/29/24", "[GEBCO](https://www.gebco.net/data_and_products/gridded_bathymetry_data/#area)",
  "Exclusive Economic Zones", "Flanders Marine Institute (2024): MarineRegions.org.", "11/29/24", "[Marine Regions](https://www.marineregions.org/eez.php)",
  "Sea Life Base Aquaculture Data", "Palomares, M.L.D. and D. Pauly. Editors. 2024. SeaLifeBase. World Wide Web electronic publication. www.sealifebase.org, version (08/2024).", "11/29/24", "[Sea Life Base](https://www.sealifebase.ca/search.php)"
)

knitr::kable(citations_df, 
             caption = "Data Citations")
```
